How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?

### 2021-10-12 analysis/paper/paper.Rmd
Edited rmd file for revision, added files for submission

---
title: "How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?"
author:
  - Gayoung Park
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2: 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  We use models from human behavioral ecology to examine stone artifacts from 23 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene when new tools, such as stemmed points and blades, appeared in the archaeological record. We focus on two questions: what changes in foragers’ landscape use were associated with the introduction of new tools? And what changes in the way people use habitation sites and mobility were associated with the new technology? To answer these questions we present quantitative analyses of artifact volumetric density, retouch frequency, composition of toolkits, and artifact raw materials. We explore the environmental and demographic contexts by applying paleoclimate simulations and summed probability distribution models. We find that quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points throughout the Late Pleistocene. Our results show that forager groups using stemmed points were associated with occupation of marginal or extreme environments. In addition, groups with stemmed points were associated with expedient technologies, indicating residential and less mobile behaviours. The environmental context of this technological innovation was a gradual decrease in temperature into the LGM. A population steadily increased before the appearance of stemmed points.

@@

The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (ca. 42-35 ka) in Korea is often assumed to have transformed forager lifeways because it was thought to have indicated more specialized hunting practices [@seong2008tanged; @chang2013human]. The stemmed point is a projectile point made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. These types of artifacts are often called tanged points in the western hemisphere [@serwatka20162d; @ivanovaite2020all], but we use ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent a hafting projection used for joining the lithic to a wooden shaft. If these artifacts were projectile weapon tips, then stemmed points represent the first appearance of hafting technology in Korea. Most open sites in Korea do not preserve organic remains due to acidic soil. There are a few limestone cave sites containing human fossils or faunal remains (e.g. Durubong and Jeommal Cave), which located in constrained area to access in North Korea or have been rarely analyzed to directly apply to other archaeological research[@bae2013paleolithic]. Therefore stone artifacts are our only window to examine hunting and subsistence strategies. Here we use a behavioral ecological framework to explore questions about the implications of the appearance of this new technology. We used data from 23 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the of the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.

@@

Previous work on this technological change in Korea has focused on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin; @seong2016eastern; @bae2017late]. To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail. Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artifact assemblages using Human Behavioral Ecology (HBE) concepts. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility; @seong2015diversity; @seong2008tanged], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that were part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.

@@

In the Eastern hemisphere, the most striking technological innovations during the Late Pleistocene are found in Northeast Asia, such as northern China, Japan and Korea. These innovations include blade technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, ground stone tools etc, and frequent use of high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean]. Previous work has argued that the appearance of new stone artifact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @bae2012nature; @bae2017origin]. For example, the new dating results from Zhoukoudian Upper Cave from northern China (35.1–33.5 ka) and material evidence such as ornaments and hominin fossil show close relation to one of multiple dispersal events out of Africa into Eurasia [@li2018re].

@@

In Japan, Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38 ka, accompanied by remains of Homo sapiens found in Okinawa Island (ca. 36 ka) [@yamaoka2012use; @izuho201421]. These technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421]. These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes and led to the megafaunal extinctions [@yamaoka2012use; @bae2017origin; @nakazawa2018quaternary; @morisaki2019early; @norton2010nature; @bae2017late; @nakazawa2017pleistocene; @kaifu2019palaeolithic]. Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [@ono2002radiocaron; @yamaoka2012use]. An important example is Ishinomoto 8-ku, which is one of the earliest sites, dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture. A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools. The main lithic raw materials are chert and andesite [@izuho201421]. Stemmed points appear in Japan much later than in Korea, dated to between about 15,500 and 13,800 cal BP [@ono2002radiocaron; @tsutsumi2007dynamics].

What makes the Korean Late Pleistocene technological transition distinctive from what we see at Shuidonggou and in Japan? The key difference is that the earliest signs of new technologies in Korean assemblages are stemmed points, followed by blade technologies. Stemmed points appeared in northern China and Japan after they appear in the Korean Peninsula. The stemmed points from the Bonggok site had been known as the oldest stemmed points dated to ca. 41.5 ka, and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin]. After blades became widespread in Korea at ca. 27 ka, and were used to make the stemmed points, the shape of points became more standardized with one or two ridges on the dorsal side and triangular cross section. Replacing flakes with blades as blanks of the stemmed points led to an increase in quantity of the points over time. In addition to the stemmed points, flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@bae2010origin; @lee2013current; @Lee_2016]. However, recently excavated Hajinri site contain stemmed points and blades dated to 42 ka, which is even older than Bonggok site.

@@

The alternative model, which we call the '*in situ*' model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the south of the Korean peninsula, with no apparent external influence [@seong2008tanged]. This claim is supported by Korea having the earliest appearance of stemmed points in Northeast Asia, at sites such as Bonggok, Hwadaeri, Hopyeongdong, and Yonghodong dated to approximately 42-35 ka BP [@seong2009emergence]. Seong(2009) presents the increased blade-to-flake ratios of lithic assemblages after the appearance of stemmed points to support the *in situ* model, based on the premise that the blade industry represents a new exogenous technology while flakes indicate a continuing local one.

## Behavioural ecology and forager land use behaviours {-}

Our brief review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focused on timing, location, origins and description of these new tools [@seong2008tanged; @seong2009emergence; @bae2010origin; @lee2013current; @Lee_2016], but modeling the land-use behaviors associated with these assemblages has rarely been undertaken. We focus on the appearance of stemmed points in Korea because these are a new technology that likely represented new hunting behaviors. This is because the hafting implies throwing techniques that can reach long-distance targets [@Keeley_1982; @kuhn2015artifacts]. We draw on behavioral ecological theory to model the effects of this technological change. Behavioral ecology theory offers structured models and testable hypotheses based on optimality assumptions derived from principles of economic rationality and environmental adaptation [@winterhalder1992evolutionary; @prentiss2019handbook]. Specifically, we use the patch choice model, which is built around the marginal value theorem, to predict that forager mobility between patches and inter-patch travel times will increase when resource patch productivity decreases [@smith1991inujjuamiunt; @smith1983anthropological; @bettinger1997evolutionary; @wolverton2015patch; @llano2015optimal]. We model individual artifact types as a kind of resource patch. Foragers may find a more costly and complicated technology (such as stemmed points, relative to flakes) optimal if they have a long cumulative time in use, if they are maintainable and reusable in a landscape where lithic raw material supply is uncertain [@torrence1983time; @bleed1986optimal]. That is, they may move to a patch where they can stay longer (i.e. an artifact type that has a long and extendible use-life) when they are not certain of the productivity and travel times of other patches [@kuhn2015artifacts].

We ask two questions to investigate the period around the introduction of stemmed points in Korea. First, what changes in foragers’ landscape use were associated with the introduction of the new tools? Specifically, was the use of stemmed points associated with occupation of marginal or extreme environments? Our assumption here is that this new technology, and the ability to reach long-distance targets, might have allowed foragers to explore and sojourn in less productive landscapes by increasing their mobility. For example, in Australia during the mid-Holocene, Hiscock et al. argue that the first appearance of microlithics in many locations represent a portable and multi-functional toolkit that minimize travel expenses and increase tool readiness to adapt to patchy and unpredictable resources associated with environmental change [@Hiscock_1994; @Hiscock_Clarkson_Mackay_2011]. Similarly, in Japan, [@Morisaki_Izuho_Terry_Sato_2015] argue that the transition from trapezoid to blades and projectile points around 25 ka in north Paleo-Honshu Island enabled the foragers to extend their occupation into cold grassland landscapes.

@@

After the first Paleolithic sites such as Dongkwanjin and Yondaebong were discovered during the mid-1930s, more than 200 Paleolithic sites have been discovered in South Korea [@norton2000current; @lee2017korean; @lee2019were]. We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change. We acknowledge the limitations of directly accepting the dating results from excavation reports since the majority of sites in Korea are open-air localities which are more prone to be disturbed compared to cave sites. Therefore, we carefully excluded sites whose assemblages are not matched with their dating results. 
This resulted in a sample of 35 assemblages from `r nrow(korean_archaeological_site_locations)` sites spanning 49-24 ka (Figure \@ref(fig:korea-map)). We identified multiple assemblages in a site where culturally sterile deposits separated artifact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artifacts. For example, the Hwadaeri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We obtained the data from published excavation reports held by provincial museums.

@@

The 35 assemblages consist of (non-blade) cores, blade cores, blades, flakes, debris, hammers, choppers, planes, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc. We excluded some artifacts from our sample because of uncertainty about their typology. The stone artifacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in the excavation reports. We have visited most of these collections to obtain permission to study the materials directly.

@@

To answer our two research questions, we compiled information about artifact density, retouch frequency, toolkit composition, raw materials, site elevation, and radiocarbon ages.

@@

Our study of raw material consumption patterns assumes that stone tool was transported and utilized to optimize mobility and minimize risks of not having artifacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic]. The representation of raw materials in an assemblage can be an indicator of site occupation patterns. For example, a large variability of raw materials in a site can suggest long-term settlement, while a small assemblage of exotic raw materials can indicate higher mobility [@valde2020models]. Here we want to know the specific raw materials that relate to the appearance of a new technology, and investigate site occupation patterns by examining changes in the proportions of raw material types in the assemblages.

@@

In addition to these data on human behavior from the stone artifact assemblages, our HBE models require environmental proxies to build hypotheses of behavioral change influenced by the surrounding environment. However, there are few paleoenvironmental proxies in Korea relevant to our study period. For example, geoarchaeological observations simply suggest that environment of the period was overall cooler and drier during MIS 3 [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late]. To supplement the limited availability of high-resolution local proxies, we used climate information from a simulated data set of paleoclimate including global monthly temperature, covering the last 120,000 years [@beyer2020high]. We extracted annual temperature values from the data generated by @beyer2020high for all site locations in our sample. In addition, we extracted site elevation values from a raster, as a proxy to represent different environments at local scales, to identify how these impact mobility and site occupation, and their relationship with the stemmed points.

@@

We used radiocarbon dates as a proxy for human population to infer the demographic context of technological change. Summed probability distributions (SPD) of radiocarbon dates are increasingly used to infer temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread]. Despite concerns about limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon; @carleton2021evaluating], several studies have shown good agreement between SPDs and other archaeological indicators (e.g. site counts, settlement size) so this method may allow a first approximation of population fluctuations [@french2015upper; @palmisano2017comparing]. Although this is often used with samples of thousands of ages, this method can also be useful for small sample sizes. For instance, @timpson2014reconstructing apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n = 243). In addition to the 93 radiocarbon ages from our targeted sites, we included all available dating results (540 ages) from other Korean Paleolithic sites that have younger and older assemblages. We compiled a total of 633 dates from 113 sites and applied a chronometric hygiene protocol to assess the reliability of radiocarbon determinations in relation to target events. Following the criteria from @napolitano2019reevaluating and @fitzpatrick2006critical, we kept ages from dating short-lived terrestrial material (i.e., plant remains) and charcoal, and having laboratory name and number. We excluded OSL dates and sites with only one date. We ended up using `r nrow(dates_clean)` radiocarbon ages from `r nrow(korean_archaeological_site_locations)` sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.

There is a debate about whether past population growth has played a role in technological change [@shennan2001demography; @hosfield2005individuals;  @kline2010population]. @shennan2009pattern observed a correlation between population decrease and technological simplification in his study of lithic arrowheads from the southern Scandinavian Mesolithic. He found that the lithics became less complex, and point shape changed as the population level dropped. By proposing the generalized null demographic model built on cultural transmission theory, @lycett2010demographic also claimed that relatively lower effective population sizes in East Asia could be strongly associated with a lower frequency of handaxe sites, a lower percentage of bifaces, and morphological difference between East Asian handaxes and classic Acheulean artifacts. They also suggested that a lack of Levallois technologies in East Asia might be similar to the previous example of handaxe. On the other hand, @collard2013population argue that regional archaeological data do not support the population-driven models. For example, @buchanan2016drivers concluded that changes in point types from \>13,000 years ago to 400 years ago in Texas were related to environmental risk, and not dependent on population size. We explored SPD to test for correlations between population, environmental proxies and technological change. We used the R package 'rcarbon' [@cremaanalysing] to perform statistical analyzes and conduct model testing using Monte Carlo methods [@timpson2014reconstructing; @crema2016summed; @palmisano2020holocene]. We generated exponential, uniform, linear, and logistic models and evaluated the model goodness-of-fit by applying Akaike's Information Criterion (AIC) as a selection criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].

@@

Quartz, quartz vein and quartzites were the most frequently found raw materials, and they were constantly used throughout the Late Pleistocene (Figure \@ref(fig:raw-materials)). For example, both Galdamri-A, dated to 44 ka with a small assemblage, and Hopyeongdong-B, dated to 27.5 ka with a larger assemblage, consist of only the quartz related materials. The use of chert, hornfels, rhyolite, and shale increased after 42 ka, but these raw materials remained a small proportion of assemblages until 25 ka. They are very suitable to make elongate blades and stemmed points because of the predictability of flaking afforded by their fine-grained texture. For example, Hajinri has the largest number of stemmed points (n=86) among the assemblages studied here, and most stemmed points from Hajinri are made from shales. Other raw materials, such as porphyry and sandstone, were found in just a few assemblages in small proportions.

>>>>

---
title: "How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?"
author:
  - Gayoung Park
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2: 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  We use models from human behavioral ecology to examine stone artifacts from 23 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene. This is an important period because new tools, such as stemmed points and blades, appeared in the archaeological record. We focus on two questions: what changes in foragers’ landscape use were associated with the introduction of new tools? And what changes in the way people use habitation sites and mobility were associated with the new technology? To answer these questions we present quantitative analyses of artifact volumetric density, retouch frequency, composition of toolkits, and artifact raw materials. We explore environmental and demographic contexts by applying paleoclimate simulations and summed probability distribution models. We find that quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points throughout the Late Pleistocene. Our results show that forager groups using stemmed points may have been associated with occupation of marginal or extreme environments. Groups with stemmed points were associated with expedient technologies, indicating residential and less mobile behaviors. The environmental context of this technological innovation was a gradual decrease in temperature into the LGM. Population increased before the appearance of stemmed points.

@@

The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (ca. 42-35 ka) in Korea is often assumed to have transformed forager lifeways because it was thought to have indicated more specialized hunting practices [@seong2008tanged; @chang2013human]. Stemmed points are projectile points made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. These types of artifacts are often called tanged points in the western hemisphere [@serwatka20162d; @ivanovaite2020all], but we use ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent a hafting projection used for joining the lithic to a wooden shaft. If these artifacts were projectile weapon tips, then stemmed points represent the first appearance of hafting technology in Korea. However, most open sites in Korea do not preserve organic remains due to acidic soil, so shafts have not been found. There are very few limestone cave sites containing human fossils or faunal remains (e.g. Durubong and Jeommal Cave), which located in constrained area to access in North Korea or have been rarely analyzed to directly apply to other archaeological research[@bae2013paleolithic]. Therefore stone artifacts such as stemmed points are our only window to examine hunting and subsistence strategies. Here we apply a behavioral ecological framework to explore questions about the implications of the appearance of this new technology. We use data from 23 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.

@@

Previous work on this technological change in Korea has focused on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin; @seong2016eastern; @bae2017late]. To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail. Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artifact assemblages using Human Behavioral Ecology (HBE) models. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility; @seong2015diversity; @seong2008tanged], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that were part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.

@@

In the Eastern hemisphere, some of the most striking technological innovations during the Late Pleistocene are found in Northeast Asia, such as northern China, Japan and Korea. These innovations include blade technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, ground stone tools etc, and frequent use of high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean]. Previous work has argued that the appearance of new stone artifact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @bae2012nature; @bae2017origin]. For example, the new dating results from Zhoukoudian Upper Cave from northern China (35.1–33.5 ka) and material evidence such as ornaments and hominin fossil show close relation to one of multiple dispersal events out of Africa into Eurasia [@li2018re].

@@

In Japan, Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38 ka, accompanied by remains of *Homo sapiens* found in Okinawa Island (ca. 36 ka) [@yamaoka2012use; @izuho201421]. These technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421]. These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes and led to the megafaunal extinctions [@yamaoka2012use; @bae2017origin; @nakazawa2018quaternary; @morisaki2019early; @norton2010nature; @bae2017late; @nakazawa2017pleistocene; @kaifu2019palaeolithic]. Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [@ono2002radiocaron; @yamaoka2012use]. An important example is Ishinomoto 8-ku, which is one of the earliest sites, dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture. A total of 500 stone artifacts, mostly chert and andesite, were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools [@izuho201421]. Stemmed points appear in Japan much later than in Korea, dated to between about 15,500 and 13,800 cal BP [@ono2002radiocaron; @tsutsumi2007dynamics].

What makes the Korean Late Pleistocene technological transition distinctive from what we see at Shuidonggou and in Japan? The key difference is that the earliest signs of new technologies in Korean assemblages are stemmed points, followed by blade technologies. This contrasts with China and Japan where stemmed points appear after their first appearance in the Korean Peninsula. Stemmed points from the Bonggok site are currently accepted as the oldest stemmed points, dated to ca. 41.5 ka, and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin]. After blades became widespread in Korea at ca. 27 ka, they were used to make the stemmed points, and the shape of points became more standardized with one or two ridges on the dorsal side and a triangular cross section. Replacing flakes with blades as blanks of the stemmed points led to an increase in quantity of the points over time. In addition to the stemmed points, flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@bae2010origin; @lee2013current; @Lee_2016]. A notable exception to this pattern is the recently excavated Hajinri site which is an outlier with both stemmed points and blades appearing together at 42 ka [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et].

@@

The alternative model, which we call the '*in situ*' model claims that stemmed points and other Late Paleolithic technologies, including blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence [@seong2008tanged]. This claim is supported by Korea having the earliest appearance of stemmed points in Northeast Asia, at sites such as Bonggok, Hwadaeri, Hopyeongdong, and Yonghodong dated to approximately 42-35 ka BP [@seong2009emergence]. Seong(2009) presents the increased blade-to-flake ratios of lithic assemblages after the appearance of stemmed points as support of the *in situ* model, based on the premise that the blade industry represents a new exogenous technology while flakes indicate a continuing local one.

## Behavioral ecology and forager land use behaviors {-}

Our brief review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focused on the timing, location, origins and description of these new tools [@seong2008tanged; @seong2009emergence; @bae2010origin; @lee2013current; @Lee_2016]. Modeling the land-use behaviors associated with these assemblages has rarely been undertaken. We focus on the appearance of stemmed points in Korea because these are a new technology that likely represented new hunting behaviors. This is because the hafting implies throwing techniques that can reach long-distance targets [@Keeley_1982; @kuhn2015artifacts]. We draw on behavioral ecological theory to model the effects of this technological change. Behavioral ecology theory is well established in international archaeology and offers structured models and testable hypotheses based on optimality assumptions derived from principles of economic rationality and environmental adaptation [@winterhalder1992evolutionary; @prentiss2019handbook]. Specifically, we use the standard patch choice model, which is built around the widely-used marginal value theorem, to predict that forager mobility between patches and inter-patch travel times will increase when resource patch productivity decreases [@smith1991inujjuamiunt; @smith1983anthropological; @bettinger1997evolutionary; @wolverton2015patch; @llano2015optimal]. We model individual artifact types as a kind of resource patch. Foragers may find a more costly and complicated technology (such as stemmed points, relative to flakes) optimal if they have a long cumulative time in use, if they are maintainable and reusable in a landscape where lithic raw material supply is uncertain [@torrence1983time; @bleed1986optimal]. That is, foragers may move to a patch where they can stay longer (i.e. prefer to make and use an artifact type that has a long and extendible use-life) when they are not certain of the productivity and travel times of other patches (i.e. the productivity and use-life of other artifact types) [@kuhn2015artifacts].

We ask two questions to investigate the period around the introduction of stemmed points in Korea. First, what changes in foragers’ landscape use were associated with the introduction of the new tools? Specifically, was the use of stemmed points associated with occupation of marginal or extreme environments? Our assumption here is that this new technology, and the ability to reach long-distance targets, might have allowed foragers to explore and sojourn in less productive landscapes by increasing their mobility. For example, in Australia during the mid-Holocene, Hiscock et al. argue that the first appearance of microlithics in many locations represent a portable and multi-functional toolkit that minimize travel expenses and increase tool readiness to adapt to patchy and unpredictable resources associated with environmental change [@Hiscock_1994; @Hiscock_Clarkson_Mackay_2011]. Similarly, in Japan, @Morisaki_Izuho_Terry_Sato_2015 argue that the transition from trapezoid to blades and projectile points around 25 ka in north Paleo-Honshu Island enabled the foragers to extend their occupation into cold grassland landscapes.

@@

After the first Paleolithic sites such as Dongkwanjin and Yondaebong were discovered during the mid-1930s, more than 200 Paleolithic sites have been discovered in South Korea [@norton2000current; @lee2017korean; @lee2019were]. We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change. It is important to note that the quality of excavation and reporting of Paleolithic sites in Korea varies greatly. Furthermore the stratigraphic integrity of the deposits is also variable, as the sites are open-air localities formed as loose accumulation of artifacts in fluviatile condition or within a colluvium developed at the edge of low-angled slope. To ensure our analysis employs reliable data, we applied a chronometric hygiene protocol (described below) to exclude sites with problematic chronologies and stratigraphy. After applying our protocol to the 200+ sites currently known, we produced a sample of 35 assemblages from `r nrow(korean_archaeological_site_locations)` sites spanning 49-24 ka (Figure \@ref(fig:korea-map)). We identified multiple assemblages in a site where culturally sterile deposits separated artifact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artifacts. For example, the Hwadaeri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We obtained the data from published excavation reports held by provincial museums.

@@

The 35 assemblages consist of (non-blade) cores, blade cores, blades, flakes, debris, hammers, choppers, planes, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc. We excluded some artifacts from our sample because of uncertainty about their typology. The stone artifacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in Korean language excavation reports. We have visited most of these collections to obtain permission to study the materials directly.

@@

To answer our two research questions, we compiled information about artifact density, retouch frequency, toolkit composition, raw materials, site elevation, and radiocarbon ages for the sites in our sample.

@@

Our study of raw material consumption patterns assumes that stone tool was transported and utilized to optimize mobility and minimize risks of not having artifacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic]. It follows that the representation of raw materials in an assemblage can be an indicator of site occupation patterns. For example, a large variability of raw materials in a site can suggest long-term settlement, while a small assemblage of exotic raw materials can indicate higher mobility [@valde2020models]. Here we quantify the specific raw materials that relate to the appearance of a new technology, and investigate site occupation patterns by examining changes in the proportions of raw material types in the assemblages.

@@

In addition to these data on human behavior from the stone artifact assemblages, our HBE models require environmental proxies to build hypotheses of behavioral change influenced by the surrounding environment. However, there are few paleoenvironmental proxies in Korea relevant to our study period. For example, geoarchaeological observations simply suggest that environment of the period was overall cooler and drier during MIS 3 [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late]. To supplement the limited availability of high-resolution local proxies, we used climate information from a simulated data set of paleoclimate including global monthly temperature, covering the last 120,000 years [@beyer2020high]. We extracted annual temperature values from the data generated by @beyer2020high for all site locations in our sample. In addition, we extracted site elevation values from a raster, as a proxy to represent environmental variation at local scales, to assist in identifying how these impact mobility and site occupation, and their relationship with the stemmed points.

@@

We used radiocarbon dates as a proxy for human population to infer the demographic context of technological change. Summed probability distributions (SPD) of radiocarbon dates are widely used by archaeologists to infer temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread]. Despite concerns about limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon; @carleton2021evaluating], several studies have shown good agreement between SPDs and other archaeological indicators (e.g. site counts, settlement size) so this method may allow a first approximation of population fluctuations [@french2015upper; @palmisano2017comparing]. Although this is often used with samples of thousands of ages, this method can also be useful for small sample sizes. For instance, @timpson2014reconstructing apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n = 243). In addition to the 93 radiocarbon ages from our targeted sites, we included all available dating results (540 ages) from other Korean Paleolithic sites that have younger and older assemblages. We compiled a total of 633 dates from 113 sites and applied a chronometric hygiene protocol to assess the reliability of radiocarbon determinations in relation to target events. Following the criteria from @napolitano2019reevaluating and @fitzpatrick2006critical, we kept ages from dating short-lived terrestrial material (i.e., plant remains) and charcoal, and having laboratory name and number. We excluded OSL dates and sites with only one date. This resulted in a sample of `r nrow(dates_clean)` radiocarbon ages from `r nrow(korean_archaeological_site_locations)` sites which we use to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.

There is a debate about whether past population growth has played a role in technological change [@shennan2001demography; @hosfield2005individuals;  @kline2010population]. @shennan2009pattern observed a correlation between population decrease and technological simplification in his study of lithic arrowheads from the southern Scandinavian Mesolithic. He found that the lithics became less complex, and point shape changed as the population level dropped. By proposing the generalized null demographic model built on cultural transmission theory, @lycett2010demographic also claimed that relatively lower effective population sizes in East Asia could be strongly associated with a lower frequency of handaxe sites, a lower percentage of bifaces, and morphological difference between East Asian handaxes and classic Acheulean artifacts. They also suggested that a lack of Levallois technologies in East Asia might be similar to the previous example of handaxe. On the other hand, @collard2013population argue that regional archaeological data do not support the population-driven models. For example, @buchanan2016drivers concluded that changes in point types from \>13,000 years ago to 400 years ago in Texas were related to environmental risk, and not dependent on population size.

We used SPD to test for correlations between population, environmental proxies and technological change. We used the R package 'rcarbon' [@cremaanalysing] to perform statistical analyzes and conduct model testing using Monte Carlo methods [@timpson2014reconstructing; @crema2016summed; @palmisano2020holocene]. We generated exponential, uniform, linear, and logistic models and evaluated the model goodness-of-fit by applying Akaike's Information Criterion (AIC) as a ranking criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].

@@

Quartz, quartz vein and quartzites were the most frequently found raw materials, and they were constantly used throughout the Late Pleistocene (Figure \@ref(fig:raw-materials)). For example, both Galdamri-A, dated to 44 ka with a small assemblage, and Hopyeongdong-B, dated to 27.5 ka with a larger assemblage, consist of only the quartz related materials. The use of chert, hornfels, rhyolite, and shale increased after 42 ka, but these raw materials remained a small proportion of assemblages until 25 ka. They are very suitable to make elongate blades and stemmed points because of the predictability of flaking afforded by their fine-grained texture. For example, Hajinri has the largest number of stemmed points (n=86) among the assemblages studied here, and most stemmed points from Hajinri are made from shale. Other raw materials, such as porphyry and sandstone, were found in just a few assemblages in small proportions.
